---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rxs2tria

<!-- badges: start -->

<!-- badges: end -->

`rxs2tria` is an `R` package to handle Quantitative Wood Anatomy (QWA) data, specifically data generated with the [ROXAS](https://roxas.wsl.ch/en/) software or its successor [ROXAS AI](https://github.com/roxas-ai/napari-roxas-ai/tree/released) *(under development)*. The package includes functionality to prepare, quality check and harmonize a QWA dataset, consisting of QWA images, corresponding ROXAS output files and accompanying metadata. The processed dataset may then be submitted to the [TRIA](https://webapps.wsl.ch/tria/) database for intra-annual and anatomical tree-ring data.

**Note:** This package is currently under development and not yet stable nor officially released.

## Installation

You can install the development version of `rxs2tria` from [GitHub](https://github.com/tria-db/rxs2tria/) with:

```{r, eval = FALSE}
pak::pak("tria-db/rxs2tria")
# or pak::pak("tria-db/rxs2tria@develop") for the develop branch
```

## Getting started

The easiest way to get started is to follow the `rxs2tria` workflow defined in the template file.
The following makes a copy of the template and saves it to your current working directory:
```{r, eval = FALSE}
library(rxs2tria)
file.copy(system.file("templates/prepare_rxs_dataset.R", package = "rxs2tria"), ".")
```


See the [vignette](https://tria-db.github.io/rxs2tria/articles/rxs2tria.html) for a detailed introduction and example workflow. You can access it locally with:
```{r, eval = FALSE}
vignette("rxs2tria")
```


## Contributions

Contributions are welcome! Please feel free to open issues or submit pull requests on the [GitHub repository](https://github.com/tria-db/rxs2tria/).


