---
title: "The ring flags Shiny app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The ring flags Shiny app}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(rxs2tria)
```

## Overview

Once cell and ring measurements have been compiled and cleaned
(with `complete_QWAdata()`, see `vignette("workflow")`), it may be
helpful to visually inspect the resulting measurement series. 
This allows you to check whether 
- the automatically estimated ring flags (wedging/missing rings, 
  radially incomplete rings, overlapping/duplicate rings) are correct;
- there are any gaps in the coverage, e.g. due to images that were 
  forgotten in the analysis process;
- there are other implausible values or outliers due to quality 
  issues (e.g. decay,   cracks, out-of-focus images) or discrete 
  anatomical features (e.g. blue or frost rings, compression wood).

The ring flags app is an interactive quality-check tool: it plots the
ring (or aggregated cell-level) measurement series for every woodpiece
in your dataset, lets you click through individual years, and records
data-quality and discrete-feature flags per ring in a table alongside
the plot. 

Note that this step is optional but we highly recommend using the app
to ensure and improve the quality of your submission.

---

## Launching the app

```{r launch}
launch_flags_app()
```

This requires a number of Suggested packages (`bslib`, `collapse`,
`htmlwidgets`, `plotly`, `reactable`, `rhandsontable`, `shiny`,
`shinyjs`), which `rxs2tria` will prompt you to install if missing.

*(screenshot: app opened with no data loaded)*

### Providing input data

Click **Load input data** in the sidebar and choose a source (from the
current R environment, or from CSV/JSON files) for up to three inputs:

- **QWA rings data** (required): the `$rings` component of a `QWAdata`
  object, i.e. `QWA_data$rings` from Step 4 in the workflow.
- **Profile data** (optional): a sector `QWAprofile` object (Step 6a
  of the workflow / see below) to allow also aggregated cell measurements 
  as plotted QWA parameters in the time series.
- **ROXAS images metadata** (optional): a `QWAimages` object, or the
  `$images` component of a `QWAmetadata` `.json` (Step 3b). This enables
  filtering by site/species, opening image files directly from the app
  (leveraging the values in the `$fname_images` column), and the Comments
  tab described below.

*(screenshot: input source modal)*

Loading new data always overwrites what's currently in the app, with a
confirmation prompt if you have unsaved edits.

#### The QWAprofiles input

The optional **Profile data** input (a `QWAprofile` object) contains cell
measurements aggregated across the ring width into *n* position bins of
equal *relative*  width (0–100% of ring width). To create such an object,
run the following (cf. `?calculate_sector_profiles`), before launching 
the flags app:

```{r, eval = FALSE}
prf_sector <- calculate_sector_profiles(
  QWA_data, 
  n_sectors = 5,
  sel_cell_params = c("la", "cwttan"),
  quant_probs = c(0.1, 0.5, 0.9)) 
# To save to file:
# write_QWAprofile(prf_sector, "path/to/output_data/example_dataset_QWAprofile_sector5.csv.gz") 
```

---

## The "Quality check series" tab

### Filtering and plot settings

The sidebar lets you subset the data (by species, site, and/or
woodpiece) and configure the main plot: which QWA parameter to show
(ring-level, or an aggregated cell-level parameter from the profile
data, with a sector selector), an optional 32-year spline detrend, the
type of mean curve, and whether to show rings already marked for
exclusion.

*(screenshot: sidebar filters and plot settings)*

### The chronology plot

The main plot shows the measurement series for the selected woodpiece
highlighted against all others in the current filter, with optional
sample-depth and ring-coverage subplots underneath. Click any point to
select that ring -- clicking a point on a *different* woodpiece's trace
switches the selection to it. Once a ring is selected, use the arrow keys
(or the arrow buttons above the ring table) to navigate to the
previous/next ring (up/down arrows) or the previous/next image (left/right
arrows). The enter/esc keys toggle the table focus. When the table is
in focus, the left/right arrows navigate across the table columns instead.

*(screenshot: chronology plot with a ring selected)*

### The ring flags table

Below the plot, a table shows every ring of the currently selected image,
with editable columns for:

- `exclude_issues` is a the most important flag: it decides whether the ring
  should be excluded from analysis. 
- Four groups of checkboxes, customisable via the table's gear icon:
  **disqualifying issues** (radially incomplete ring, wedging/missing
  ring, crossdating, compression wood, orientation), **technical issues**
  (out of focus, cracks, paraffin, compressed/overlapping/broken cells,
  tangentially incomplete, irregular thickness, air bubbles), **other
  issues** (tyloses, decay, other), and **discrete features** (blue/frost/
  light ring, intra-annual density fluctuations, traumatic resin ducts,
  trabeculae, blue/red cells, other).
- `affected_tissue` (`all`/`ew`/`lw`) to further specify which part of the 
  ring is affected by the selected flags.
- A free-text `comment` field.

In addition, there is the read-only column `duplicate_ring`, which indicates
that the same year is also present in at least one other image. The 
corresponding `exclude_dupl` column defines for each group of overlapping 
years which one is the selected best. Specifically, for each distinct
`(woodpiece_label, year)`, exactly one `image_label` has `exclude_dupl=FALSE`,
all others are set to `TRUE`. The default selected one is the ring with the
highest cell count, but you can change this here. Editing `exclude_dupl` for
one ring automatically applies the corresponding change to the other overlapping 
image(s).

Click **Update excluded in plot** to refresh the plot markers after editing 
exclusion flags.

*(screenshot: ring flags table for a selected image)*

If image metadata was loaded, you can also open the underlying image
file directly (preferring an annotated version if one exists) via the
image icon next to the selected image name.

---

## The "Comments" tab

This tab only appears if the loaded image metadata included non-empty
`comment` values (e.g. notes added during ROXAS AI processing, or
manually in the metadata app's Start tab table). It lists every
commented image with a "handled" checkbox and a button to jump straight
to that image/ring in the main tab, plus a badge showing how many
comments are still unhandled -- a convenient checklist to make sure
every flagged image was actually looked at during the cleaning process.

*(screenshot: comments overview table)*

---

## Saving your work

Click **Save ring flags** (or configure the destination first via the
gear icon next to it) to write your edits to a local CSV file and/or an
R environment variable. If a Comments tab was shown, its "handled"
status is saved alongside it, to the images metadata file/variable, so
you (or a colleague) can resume exactly where you left off.

*(screenshot: save settings modal)*

---

## Output and how it fits into the workflow

The saved rings data is the `$rings` component of a `QWAdata` object,
with the manually reviewed flag columns added or updated. Read it back
and merge it into your `QWAdata` object to finish Step 6b of
`vignette("workflow")`:

```{r merge-back}
QWA_data$rings <- read_QWAdata(
  file_rings = "path/to/output_data/example_dataset_edited_QWArings.csv")$rings

write_QWAdata(QWA_data, dir = path_out, dataset_name = dataset_name)
```

If reviewing the data in the app reveals that one or more images need to
be re-analysed in ROXAS/ROXAS AI, use `update_QWAimages()` and
`update_QWAdata()` (Step 6c) to splice in the new results rather than
reprocessing the whole dataset -- by default this resets any manual
flags for the updated images, since they may no longer apply to the
re-analysed data.