---
title: "The metadata Shiny app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The metadata Shiny app}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(rxs2tria)
```

## Overview

A TRIA submission needs metadata at every level of the sampling
hierarchy -- site, tree, woodpiece, slide -- plus dataset-level
information such as authors, funding, licensing, and related
publications (see `vignette("QWAmetadata")` for the full list of
components). Filling all of this in by hand, as data frames or JSON, is
tedious and error-prone.

The metadata Shiny app is a guided, form-based alternative: it takes the
`QWAimages` object you already built (Step 3 of `vignette("workflow")`)
as its starting point and walks you through the remaining components tab
by tab. Along the way it offers lookup tools for organisations
([ROR](https://ror.org/)), authors
([ORCID](https://orcid.org/)) and related publications/datasets
([Crossref](https://www.crossref.org/) DOIs), and validates your entries
as you go. The output is a `QWAmetadata` object, exported as a `.json`
file -- one of the three files required for every TRIA submission.

The app is entirely self-contained: it does not read or write anything
outside of what you explicitly load or export, and any partially
completed submission can be saved and reopened later.

---

## Launching the app

```{r launch}
launch_metadata_app()
```

This opens the app in your system's default web browser (or the
RStudio/Positron viewer pane). It requires a number of Suggested
packages (`bslib`, `data.tree`, `DT`, `htmlwidgets`, `httr`, `leaflet`,
`rcrossref`, `rhandsontable`, `sf`, `shiny`, `shinyjs`, `shinyTree`,
`shinyvalidate`) which are only needed for this app -- `rxs2tria` will
prompt you to install any that are missing.

*(screenshot: app opened on the Start tab)*

---

## Walking through the tabs

The app has four tabs, meant to be completed in order (a "Next"/"Previous"
button pair in each sidebar moves between them), though you can jump
back to fix earlier entries at any time.

### 1. Start

Load your input data using the **Load input data** button in the
sidebar, from one of three sources:

- **From local R environment**: point it at a `QWAimages` or
  `QWAmetadata` object already in your R session (e.g. `rxs_images` from
  Step 3 of `vignette("workflow")`).
- **From file**: a `QWAimages` `.csv` file, or a `QWAmetadata`
  `.json` file -- use the latter to resume a partially completed
  submission.
- **Load example data**: loads the small example dataset bundled with
  the package, useful for trying out the app without your own data.

*(screenshot: input source modal)*

Once loaded, the main panel shows:

- **Data structure**: an interactive tree (site > tree > woodpiece >
  slide > image) visualising the hierarchy inferred from your image file
  names, with the labels `rxs2tria` generated for each level.
- **Metadata extracted from files**: a table of everything extracted
  from the ROXAS/ROXAS AI settings and image files. Most columns are
  read-only; `band_width`, `only_ew` and `comment` can be edited
  directly in the table (which supports Excel-like copy/paste and
  autofill).

*(screenshot: data structure tree and metadata table)*

**Check this tab carefully** -- it is the foundation for everything
that follows. If the structure or labels look wrong, revisit Step 2
(`extract_data_structure()`) in `vignette("workflow")` and start over.

### 2. General

Dataset- and author-level information, in three parts:

- **Dataset information**: title, description, access rights (public or
  restricted, with an embargo date), license (CC licenses preferred), 
  and acknowledgements.
- **Authors & Funding**: two editable tables for authors and their 
  affiliations as well as funding sources. Two search tools for ROR
  and ORCID data can help you find the relevant information.
- **Related resources**: a table of related publications or datasets,
  with a Crossref-backed search by DOI, citation text, or TRIA dataset
  ID.

*(screenshot: author table with ORCID search results)*

A **validation check summary** in the sidebar lists any issues with the
entries so far (e.g. missing required fields).

### 3. Samples

Site-, tree-, woodpiece- and slide-level metadata, in one table per
level. Rows are pre-filled with the IDs and labels inferred from the
`QWAimages` or (if available) taken directly from the `QWAmetadata` 
provided as input in the Start tab. The site table also drives an 
interactive map: any site with valid coordinates gets a marker, which is
a quick way to spot mistyped coordinates.

*(screenshot: site map with markers)*

### 4. Summary

A combined validation check summary across all tabs -- resolve (or
consciously accept) any remaining issues here. Clicking **Export data
to file** downloads the completed `QWAmetadata` object as a `.json`
file.

*(screenshot: summary tab with validation table)*

---

## Output and how it fits into the workflow

The exported `.json` file is a `QWAmetadata` object -- see
`vignette("QWAmetadata")` for its structure and for how to inspect,
check, or edit it further in R. Read it back with:

```{r read-back}
QWA_meta <- read_QWAmetadata("path/to/output_data/example_dataset_QWAmetadata.json")
```

Within the overall data preparation workflow (`vignette("workflow")`),
this app covers Step 3b, run after building the `QWAimages` object in
Step 3 and before (or alongside) processing the measurement data in
Steps 4-5. The resulting `.json` file is, together with the two
`QWAdata` `.csv` files, one of the three files required for every TRIA
submission (`vignette("submission")`).

Because partially completed submissions can be saved and reopened at
any time, you do not need to finish in one sitting -- export what you
have, and load the `.json` back into the Start tab later to continue.
