---
title: "The QWAdata class"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The QWAdata class}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(rxs2tria)
```

## Overview

A `QWAdata` object is a named list with two components:

- **`$cells`** — cell-level measurements, one row per cell.
- **`$rings`** — ring-level measurements and quality flags, one row per
  annual ring per image.

At least one component must be present; the other may be `NULL`. The
object does not carry ROXAS settings or hierarchical metadata — those
stay in the accompanying `QWAimages` or `QWAmetadata` object, linked to
`QWAdata` via the `image_label` key.

---

## Building a QWAdata object

In normal use, `QWAdata` is built incrementally in three steps:

```{r build}
# Step 1: read raw ROXAS output into a QWAdata object
QWA_data <- collect_raw_data(rxs_images)

# Step 2: add derived measures and ring quality flags
QWA_data <- complete_QWAdata(QWA_data, meta = rxs_images,
                              exclude_mode = "either")

# Step 3: validate
check_QWAdata(QWA_data, meta = rxs_images)
```

`collect_raw_data()` reads the cells and rings tables from all ROXAS
output files referenced in a `QWAimages` object, cleans invalid values
(negatives, error codes → `NA`), and returns a minimal `QWAdata` with
raw measurements only.

`complete_QWAdata()` then runs the full preprocessing pipeline (see
[Completing a QWAdata object] below) and `check_QWAdata()` validates the
result.

You can also construct a `QWAdata` object directly from existing data
frames, for example after importing from another source:

```{r construct-direct}
QWA_data <- QWAdata(cells = df_cells, rings = df_rings)
```

The constructor enforces minimal column requirements (`image_label`,
`year`, `xpix`, `ypix` in cells; `image_label`, `year` in rings),
coerces column types, fills any year gaps in the rings sequence, and
checks for dating problems (duplicate years, future years). It warns if
`cwttan` is all-`NA` for any image, which indicates missing cell wall
thickness estimates — expected for conifers, acceptable for angiosperms.

---

## What the object contains

```{r inspect}
QWA_data          # same as print(QWA_data): compact overview
summary(QWA_data) # identical output
```

`print()` reports which components are present and their dimensions,
the number of woodpieces, slides, and images, the year range, and — once
the flag columns have been added — a count of incomplete, missing, and
duplicate rings.

### `$cells`

One row per cell. Columns fall into three groups:

**Identifiers**: `image_label` (links to `QWAimages`), `year`, and
pixel coordinates `xpix` / `ypix`.

**Raw measurements**: cell dimensions and wall thickness directly from
ROXAS — lumen dimensions (`drad`, `dtan`), radial distribution
(`rraddistr`), lumen area (`la`), cell wall areas (`cwa`), and cell wall
thicknesses (`cwtrad`, `cwttan`, `cwtall`, `rtsr`).

**Derived measures** (added by `complete_QWAdata()`):  `tca`, `rwd2`,
`dcwt`, `cwtall.adj`, `cdrad`, `cdtan`, `cdratio`, `sector100`, `ew_lw`
(earlywood / latewood classification based on Mork index).

### `$rings`

One row per ring per image. Columns fall into three groups:

**Identifiers**: `image_label`, `year`, `woodpiece_label`,
`slide_label`, and `cno` (cell count per ring, derived from `$cells`).

**Measurements**: mean ring width (`mrw`), ring area (`ra`), and after
`complete_QWAdata()`, earlywood and latewood widths (`eww`, `lww`).

**Quality flags** (added by `complete_QWAdata()`): logical columns
described in the next section.

---

## Completing a QWAdata object

```{r complete}
QWA_data <- complete_QWAdata(QWA_data, meta = rxs_images,
                              exclude_mode = "either")
```

`complete_QWAdata()` runs three operations in sequence and returns a new
`QWAdata` object:

1. **Derived cell measures** — computes the derived columns listed above
   that are not yet present in `$cells`. Already-present columns are
   skipped, so re-running is safe.

2. **Year sequence completion** — extends `$rings` to cover every year
   in the cells data (filling gaps with `cno = 0` and `FALSE` for flag
   columns), and checks that no ring year exceeds `outmost_year` from
   `meta`.

3. **Ring quality flags** — adds five logical columns to `$rings`:

   | Flag | Description |
   |---|---|
   | `incomplete_ring` | Ring at the inner or outer image border; MRW unreliable or unavailable |
   | `missing_ring` | No cells detected; ring added manually during crossdating (e.g. wedging ring) |
   | `duplicate_ring` | Year appears in more than one image due to overlapping slides |
   | `exclude_dupl` | `TRUE` for the duplicate ring(s) not selected as the preferred source |
   | `exclude_issues` | Convenience flag combining incomplete and/or missing rings; controlled by `exclude_mode` |

   For the `exclude_mode` options (`"either"` or `"incomplete_only"`),
   see the main `rxs2tria` vignette.

If you need finer control — for example to add derived measures without
touching the flags, or to recompute flags after manual edits — the
underlying functions `complete_measures()` and `complete_flags()` can be
called independently.

---

## Checking a QWAdata object

```{r check}
check_QWAdata(QWA_data, meta = rxs_images)
```

`check_QWAdata()` returns `TRUE` invisibly and issues warnings for any
problems found:

- **Missing CWT** (warning): any image where `cwttan` is entirely `NA`.
- **Dating problems** (warning): undated years, duplicates, gaps, future
  years, or years beyond `outmost_year` (the last only if `meta` is
  supplied).
- **Structural errors** (error): both components `NULL`, or required
  identifier columns missing.

The `meta` argument is optional but recommended: without it, the
`outmost_year` check is skipped.

---

## Persisting to disk

`QWAdata` is written as two separate CSV files — one for cells, one for
rings. Provide either a directory (files are named automatically) or
explicit paths:

```{r write-read}
# Using a directory — files named {dataset_name}_QWAdata_cells.csv.gz and
# {dataset_name}_QWAdata_rings.csv.gz
write_QWAdata(QWA_data, dir = "output/", dataset_name = "my_dataset")

# Or with explicit paths:
write_QWAdata(QWA_data,
              file_cells = "output/my_dataset_QWAdata_cells.csv.gz",
              file_rings = "output/my_dataset_QWAdata_rings.csv.gz")

# Read back:
QWA_data <- read_QWAdata(dir = "output/", dataset_name = "my_dataset")

# Read only rings (avoids loading a large cells file):
QWA_data_rings <- read_QWAdata(dir = "output/", dataset_name = "my_dataset",
                                components = "rings")
```

Compression defaults to `TRUE` for `write_QWAdata()` (unlike
`write_QWAimages()`, where it defaults to `FALSE`). `read_QWAdata()`
accepts both `.csv` and `.csv.gz`.
