---
title: "The QWAmetadata class"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The QWAmetadata class}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(rxs2tria)
```

## Overview

A `QWAmetadata` object is a named list that bundles all metadata
required for a TRIA database submission. It has ten components, one per
level of the QWA sampling hierarchy:

| Component | Description |
|---|---|
| `$images` | Image-level metadata; a `QWAimages` object. **Required.** |
| `$sites` | Site-level metadata (location, coordinates, climate) |
| `$trees` | Tree-level metadata (species, DBH, age) |
| `$woodpieces` | Woodpiece-level metadata (type: disc, core, etc.) |
| `$slides` | Slide-level metadata (preparation method, staining) |
| `$dataset` | Dataset name, description, license, embargo date |
| `$authors` | Author names, affiliations, contact information |
| `$funding` | Funding sources and grant numbers |
| `$related` | Related publications or datasets (DOIs, URLs) |
| `$resources` | Files to include in the submission |

`$images` is the only required component. All others are optional and
can be populated incrementally. The hierarchical components (`$sites`,
`$trees`, `$woodpieces`, `$slides`) are linked to `$images` via the
shared label columns (`site_label`, `tree_label`, `woodpiece_label`,
`slide_label`).

---

## Building a QWAmetadata object

In normal use, `QWAmetadata` is assembled through the interactive
metadata Shiny app, which takes a `QWAimages` object as its starting
point and guides you through all components. The output is exported as a
JSON file that can be read back at any time:

```{r build-app}
launch_metadata_app()

QWA_metadata <- read_QWAmetadata("output/my_dataset_QWAmetadata.json")
```

You can also construct a `QWAmetadata` object directly, passing any
components you already have as data frames:

```{r build-direct}
QWA_metadata <- QWAmetadata(
  images = rxs_images,      # required QWAimages object
  sites  = df_sites,        # optional data frames
  trees  = df_trees
)
```

Or from a named list, for example after reading from a custom source:

```{r build-list}
QWA_metadata <- as_QWAmetadata(my_list)
```

`as_QWAmetadata()` matches list names against the known component names,
drops any unrecognised names with a warning, and calls `QWAmetadata()`
on the remainder.

The constructor validates and aligns each supplied component against its
JSON schema: column names are cleaned, types are coerced, and
unrecognised columns are dropped. Missing required columns are added as
`NA` with a warning.

---

## Inspecting a QWAmetadata object

```{r inspect}
QWA_metadata          # same as print(QWA_metadata)
summary(QWA_metadata) # identical output
```

`print()` shows the image-level coverage (images, slides, woodpieces,
trees, sites, species, year range) and the status of each optional
component — present with row count, or flagged as not provided.

Individual components are standard data frames and can be accessed and
edited directly:

```{r access}
QWA_metadata$sites
QWA_metadata$authors

# Edit in place:
QWA_metadata$dataset$license <- "CC BY 4.0"
```

---

## Checking a QWAmetadata object

```{r check}
check_QWAmetadata(QWA_metadata)
```

`check_QWAmetadata()` validates each component in turn and returns
`TRUE` invisibly when done:

- **Absent components** are skipped with an informational message.
- **Schema compliance** (warning): checks that required columns are
  present and values are within expected ranges for each component.
- **Missing optional columns** (warning): reports optional columns not
  yet present.
- **Hierarchy consistency** (error): for `$sites`, `$trees`,
  `$woodpieces`, and `$slides`, checks that every ID present in
  `$images` has exactly one matching row, with no extras and no
  duplicates. Also verifies that derived counts (e.g. number of trees
  per site) match between the component table and `$images`.
- **`$images` validity** (delegates to `check_QWAimages()`).

---

## Completing a QWAmetadata object

```{r complete}
QWA_metadata <- complete_QWAmetadata(QWA_metadata)
```

Returns a new `QWAmetadata` object with all optional columns added and
existing values preserved:

- **`$images`**: delegates to `complete_QWAimages()`.
- **Hierarchical tables** (`$sites`, `$trees`, `$woodpieces`,
  `$slides`): if a table is absent, it is pre-filled with the IDs and
  counts derived from `$images` (the same starting point the Shiny app
  uses), with all other columns initialized as `NA`. If a table is
  already present, only missing optional columns are added.
- **Flat tables** (`$dataset`, `$authors`, `$funding`, `$related`): if
  absent, a one-row skeleton with all schema columns (all `NA`) is
  created so you can see the expected fields. If already present, only
  missing optional columns are added.
- **`$resources`**: not modified by `complete_QWAmetadata()`; use
  `add_resources()` to populate this component.

---

## Adding resources

Resources (data files to be included in the submission) are managed
separately via `add_resources()`, which scans a directory, infers the
resource type of each file from its name, and appends the result to
`$resources`:

```{r resources}
QWA_metadata <- add_resources(QWA_metadata, path = "output/submission_files/")
```

---

## Persisting to disk

`QWAmetadata` is saved as a JSON file, which preserves the nested list
structure and all component tables in a single file:

```{r write-read}
write_QWAmetadata(QWA_metadata, "output/my_dataset_QWAmetadata.json")

QWA_metadata <- read_QWAmetadata("output/my_dataset_QWAmetadata.json")
```

Partial submissions can be saved and reloaded at any time. The JSON
format is human-readable, so components can also be inspected or edited
outside of R if needed.
