---
title: "Adding supplementary resources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding supplementary resources}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(rxs2tria)
```

## Overview

A TRIA submission must include at minimum three files: the `QWAmetadata`
`.json` and the two `QWAdata` `.csv` files (cells and rings). Beyond this
minimum, you may optionally submit **supplementary resources** — the
original and annotated images, the raw ROXAS / ROXAS AI output files,
reference series, and so on — so that others can trace, reproduce, or
build on your analysis.

Supplementary resources are described in the `$resources` component of
the `QWAmetadata` object. This component is a table (a
[tibble][tibble::tibble]) with **one row per file**. It is a *manifest*:
it records what each file is, where it sits in the data hierarchy, and
whether it is actually ready to be submitted. Building it does **not**
move, rename, or compress any files — they stay where they are on your
machine and are referenced by path.

You build the table with two functions:

- `add_resources()` — the convenient entry point; scans a directory and
  updates the `$resources` component of a `QWAmetadata` object in place.
- `compile_resources()` — the underlying worker; returns a plain
  resources table and is useful when you want to inspect or manipulate
  it directly before attaching it.

As described in `vignette("submission")`, a submission's supplementary
files are expected to live in a **single directory** (which you zip up
before uploading) — the functions scan it recursively, so you may use
your preferred nested file structure within that directory. Both
functions also accept a path to an already-zipped (non-nested) archive
directly, instead of a directory — see [Re-checking an existing table or
archive] below.

---

## The resources table

Each row has the following columns:

| Column | Meaning |
|---|---|
| `resource_name` | Path of the file relative to the directory you scanned (not just its basename) — this is also the path the file will have inside the zip archive. |
| `resource_type` | The kind of file, inferred from its name (see below). |
| `linked_level` | Level of the data hierarchy the file pertains to: `"dataset"`, `"woodpiece"`, `"slide"`, `"image"`, or `"analysis"`. |
| `linked_label` | Which entity at that level the file belongs to (e.g. a specific `image_label`). Auto-filled for `image`/`analysis`/`slide`/`woodpiece` levels where possible, otherwise `NA`. |
| `fname_resource` | Absolute path to the file on your machine, or `NA` if it was found by scanning a zip archive directly (will always be removed before upload to TRIA). |
| `status` | `"ok"`, `"review"`, or `"ignore"` — see [Readiness status] below. |
| `note` | Explanation for `status` (why a file needs review, or is ignored). |

Every file found in the directory gets a row — nothing is silently
dropped — so the table is a complete, self-explanatory record of what
you're about to submit. `resource_name` includes the subdirectory path
(e.g. `"tree1/slide2/IMG1_Output_Cells.txt"`) rather than just the
basename, so two files with the same name in different subfolders stay
distinguishable.

---

## Basic workflow

Start from a `QWAmetadata` object — either one you built earlier in the
preparation workflow, or one read back from disk:

```{r read-meta}
QWA_meta <- read_QWAmetadata("output_data/my_dataset_QWAmetadata.json")
```

Point `add_resources()` at the directory holding your supplementary
files (subdirectories are included):

```{r add-basic}
QWA_meta <- add_resources(QWA_meta, path = "submission_files")
```

Inspect the result — the console already told you if anything needs
review (see [Readiness status]). Complete any resource types or labels
that require manual edits (see [Manual editing]), then write the object
back to disk:

```{r inspect-write}
QWA_meta$resources

write_QWAmetadata(QWA_meta, "output_data/my_dataset_QWAmetadata.json")
```

### Working with the table directly

If you prefer to assemble and edit the table before attaching it, use
`compile_resources()`. It takes the same arguments:

```{r collect-direct}
res <- compile_resources("submission_files", rxs_images = QWA_meta$images)

# edit `res` as needed, then attach it:
QWA_meta$resources <- res
```

`rxs_images` (a [QWAimages] object) is required: its `roxas_version`
attribute disambiguates resource types that classic ROXAS and ROXAS AI
name identically (see below), and its labels auto-populate `linked_label`
(see [Linking resources to the data hierarchy]).

---

## Re-checking an existing table or archive

Both functions accept a previous resources table (`res` for
`compile_resources()`; automatic via `x$resources` for `add_resources()`)
to re-check against, and `path` can point to a zip archive instead of a
directory — useful to re-verify an already-packaged submission zip is
still complete right before uploading it:

```{r recheck-zip}
QWA_meta <- add_resources(QWA_meta, path = "submission_files.zip")
```

When a previous table is available, it's treated as ground truth rather
than re-inferred from scratch:

- A file listed in the table **and** still found in `path` keeps its
  `resource_type`/`linked_level`/`linked_label`/`description` exactly as
  they are — any manual edits you made survive the re-run.
- A file the table lists (and hasn't marked `"ignore"`) that is **not**
  found in `path` aborts the call: the table no longer matches what's
  actually there, and either needs fixing or the file needs to be put
  back.
- A file found in `path` that the table doesn't know about is, by
  default, left out — the table is trusted as the complete list, and you
  just get a console count of how many were skipped. Pass
  `add_new_files = TRUE` (on either function) to pick these up instead,
  freshly typed, e.g. right after you've added new files to the
  directory:

```{r recheck-add-new}
QWA_meta <- add_resources(QWA_meta, path = "submission_files",
                           add_new_files = TRUE)
```

Because an untracked file is always treated as "not yet seen" rather
than "deliberately excluded," simply deleting its row from the table
does not permanently exclude it — a later re-run (with `add_new_files =
TRUE`) would just pick it up again. To permanently exclude a real file
without removing it from `path`, set its `resource_type` to `"junk"`
instead of deleting its row (see [Readiness status] below) — that stays
stable across re-runs for as long as the file exists in `path`.

---

## Readiness status

Every row is assigned a `status` by [check_resources()] (run
automatically by `compile_resources()`/`add_resources()`, and re-runnable
by hand after you edit the table):

- **`"ok"`** — recognised, passes every check, ready to submit as-is.
  An {rxs2tria}-generated file found among your supplementary files
  (`QWAmetadata.json`, the `QWAcells`/`QWArings`/`QWAprofile`
  `.csv(.gz)` files) also gets `"ok"` here — you may include it in the 
  supplementary files zip or submit it individually, either is fine.
- **`"review"`** — needs your attention before this dataset can be
  zipped for submission. `note` says why: an unrecognised (`"other"`)
  file without a `description` (see below), a missing `linked_label`
  where one is expected, more than one file of a type expected to occur
  once per level, a resource type specific to the other ROXAS version,
  or a `linked_label` that doesn't match any known label.
- **`"ignore"`** — a junk file (ROXAS `_bu` backups, stray
  calibration/preview files, `Thumbs.db`, `desktop.ini`, `.DS_Store`,
  Office lock files, ...). Never part of what gets zipped as
  supplementary resources. You may want to remove these files from the
  directory yourself to decrease the submission file size. But if you
  don't remove them, they will in any case be ignored when we upload the
  dataset to TRIA. If a file doesn't match any built-in junk pattern but
  you'd still like it permanently excluded, set its `resource_type` to
  `"junk"` yourself — it gets the same treatment, and (unlike deleting
  its row) stays excluded across re-runs, see [Re-checking an existing
  table or archive].

The console message tells you where things stand: once no row needs
review, you'll see a confirmation that the dataset is ready to zip;
otherwise the files that need attention are listed together with their
`note`. A file typed `"other"` is treated as ready (`"ok"`) as soon as
you give it a `description` explaining why it belongs in the submission
as-is (see [Adding an optional description]) — you don't need to also
assign it a "real" `resource_type` unless one actually applies.

```{r fix-and-recheck}
res$resource_type[res$resource_name == "tree1/slide2/odd_name.tif"] <- "image_original"
res$linked_level[res$resource_name == "tree1/slide2/odd_name.tif"] <- "image"

# re-run the checks after manually editing the table:
res <- check_resources(res, QWA_meta$images)
```

---

## Resource types

`resource_type` is inferred from each file name by matching it against a
table of naming patterns bundled with the package
(`system.file("extdata", "resource_types.csv", package = "rxs2tria")`).
This table is the single source of truth for pattern matching, so it is
reproduced here directly rather than duplicated by hand:

```{r resource-types-table, eval = TRUE, echo = FALSE}
tbl <- vroom::vroom(
  system.file("extdata", "resource_types.csv", package = "rxs2tria"),
  delim = ",", show_col_types = FALSE, progress = FALSE
)
tbl <- tbl[, c("resource_type", "linked_level", "example_names",
 "identifier", "source")]
knitr::kable(tbl)
```

Patterns are tested most specific first and the first match wins. 
Where classic ROXAS and ROXAS AI produce identically
named files (e.g. `*_annotated.jpg`), the `roxas_version` attribute of
`rxs_images` determines which of the two resource types applies —
patterns whose `source` column is `"roxas"` or `"roxas_ai"` are only
considered when they match it.

**Anything else** is typed as `"other"` (with `linked_level = NA` and
`status = "review"`).


## Manual editing

### Correcting an inferred type

Inference is name-based, so an unconventional file name may be typed as
`"other"` or mis-typed. Edit the table directly to fix it (matching on
the full `resource_name`, subdirectory included — see [Readiness
status] for re-running the checks afterwards):

```{r fix-type}
res$resource_type[res$resource_name == "tree1/slide2/odd_name.tif"] <- "image_original"
res$linked_level[res$resource_name == "tree1/slide2/odd_name.tif"] <- "image"
```

### Linking resources to the data hierarchy

`linked_level` says *what kind* of entity a file belongs to;
`linked_label` says *which one*. For `image`-, `analysis`-, `slide`-, and
`woodpiece`-level files, the label is filled automatically from
`rxs_images` — the `$images` component of your `QWAmetadata` object (a
[QWAimages] object).

`add_resources()` does this for you: it passes `x$images` as `rxs_images`,
so resources are matched to their label without any extra step. The
console reports how many were matched, e.g. `Matched 42/45 resources to a
data-structure label`.

For `image`/`analysis` files, matching works by testing whether a
resource's *basename* (`resource_name` with any subdirectory stripped)
**starts with** an image's base name (`org_img_name` with its extension
stripped); the longest match wins, so labels that share a prefix are not
confused (e.g. `S22_L1` vs `S22_L10`).

For `slide`/`woodpiece` files (e.g. `SLIDENAME_Preview.jpg`,
`WPNAME_Panorama.tif`), there is no "original" slide/woodpiece name
stored anywhere to match against directly — only the fully constructed
`slide_label`/`woodpiece_label`. Instead, the original identifier is
recovered from the images that share a `slide_label` (or
`woodpiece_label`): since sibling images are almost always named with a
common prefix followed by an image-specific suffix, the longest common
prefix of their `org_img_name` values recovers that original identifier,
whatever the naming convention. A slide/woodpiece with only a single
image has no sibling to compare against, so its trailing token (the part
after the last separator, e.g. `_1` in `SITEA_PISY_01_2_1`) is stripped
instead, on the assumption that it identifies the image rather than the
slide/woodpiece — if that assumption doesn't hold for a particular file,
fill the label in manually.

Files that do not match anything are left as `NA` — which is why
resource types that normally need a `linked_label` are flagged
`status = "review"` until one is filled in (see [Readiness status]).

With `compile_resources()`, `rxs_images` is a required argument:

```{r collect-structure}
res <- compile_resources("submission_files", rxs_images = QWA_meta$images)
```

`dataset`-level resources are never inferred — a dataset-level file
applies to the whole submission, so there is no single entity to match it
to. Fill these in only when a file happens to apply to something more
specific than the whole dataset (e.g. a reference series tied to one
woodpiece rather than the dataset):

```{r fill-label}
res$linked_label[res$resource_name == "reference_series/SITEA_PISY_01.rwl"] <-
  "SITEA_PISY_01"
```

A `linked_label` of `NA` is acceptable for types that don't require one —
it simply means the file is not tied to a specific entity below its
level (a dataset-wide reference chronology is a natural example). Always
review the auto-filled labels, too: the prefix-based matching is a
heuristic and can occasionally miss or mismatch unconventional file
names.

---

### Adding an optional description

The resources schema also allows a free-text `description` per file,
useful when a resource needs context that its type does not convey — or
to justify why an unrecognised (`"other"`) file should be submitted
as-is. It is not created by `compile_resources()`; add it yourself when
helpful:

```{r description}
res$description[res$resource_type == "reference_series"] <-
  "Master chronology used for crossdating, EPS > 0.85."
```

---

## Where this fits in the workflow

Compiling resources is the optional final step of the preparation
workflow (see `vignette("workflow")`, Step 7). Once `$resources` is
populated and the `QWAmetadata` object is written to disk, the manifest
travels with the submission. See `vignette("submission")` for how to
package the required and supplementary files together and what happens
to them after upload.
