---
title: "rxs2tria"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rxs2tria}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

rxs2tria is an R package to handle Quantitative Wood Anatomy (QWA) data, specifically data generated with the [ROXAS](https://roxas.wsl.ch/en/) or [ROXAS AI](https://github.com/roxas-ai/napari-roxas-ai/tree/released) software. The package includes functionality to prepare, quality check and harmonize a QWA dataset, consisting of QWA images, corresponding cell and ring measurement files and accompanying metadata. The processed dataset may then be submitted to the [TRIA](https://webapps.wsl.ch/tria/) database for intra-annual and anatomical tree-ring data.

**Note:** This package is currently under development and not yet stable nor officially released.

# Installation

The package is (not yet) on CRAN. You can install `rxs2tria` from [GitHub](https://github.com/tria-db/rxs2tria/):

```{r, eval = FALSE}
pak::pak("tria-db/rxs2tria")

# or with another remote package installer, such as:
# devtools::install_github("tria-db/rxs2tria")
# remotes::install_githbu("tria-db/rxs2tria")
```

## Development version

To get the newest features and bug fixes, you can install the development version:
```{r, eval = FALSE}
pak::pak("tria-db/rxs2tria@develop")
```

## Troubleshooting
Ensure that the installer libarary (pak/devtools/remotes) is properly installed. 
On Windows OS, you may need to install Rtools to build packages from source.
Run `pkgbuild::check_build_tool(debug=TRUE)` to check for build tools, and restart RStudio/Positron after the installation is completed.


# Quick start

The easiest way to get started is to copy the rxs2tria workflow template to your
working directory and fill in your dataset-specific paths and settings, 
then run through each step.

The following creates a file `prepare_rxs_dataset.R` in your current directory. 
```{r, eval = FALSE}
library(rxs2tria)
file.copy(system.file("templates/prepare_rxs_dataset.R",
                      package = "rxs2tria"), ".")
```

Open the copied file and follow the instructions therein, or see the
[workflow vignette](workflow.html) for more details.