---
title: "rxs2tria"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rxs2tria}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction to rxs2tria

**Note:** This package is currently under development and not yet stable
nor officially released.

## Installation

You can install the development version of `rxs2tria` from
[GitHub](https://github.com/tria-db/rxs2tria/) with:

```{r, eval = FALSE}
pak::pak("tria-db/rxs2tria") # or pak::pak("tria-db/rxs2tria@develop") for the even more experimental develop branch
```

## Reading in ROXAS data

Start by attaching the package:

```{r setup, eval = FALSE}
library(rxs2tria)
```

Set the path to a folder containing the ROXAS output files of your QWA
dataset. The files may be organized in subdirectories. The
`get_roxas_files()` function checks for the presence of the following
required files and returns their full filepaths. - ROXAS cells output
files (named `{IMAGEID}_Output_Cells.txt`) - ROXAS rings output files
(named `{IMAGEID}_Output_Rings.txt`) - ROXAS settings files (named
`{IMAGEID}_ROXAS_Settings.txt`) - corresponding JPEG image files
(`{IMAGEID}.jpg` or `{IMAGEID}.jpeg`)

```{r files, eval = FALSE}
path_in <- '/Users/maranaegelin/Documents/QWAdata/LTAL_S22/rxs_out' #'./path/to/ROXAS_output_data'
files <- get_roxas_files(path_in)
str(files)
```

### Extracting the data structure

QWA data are organized hierarchically, i.e., each analyzed *image* was
taken and potentially cropped out from a *slide* with the mounted
microsection, each microsection was cut from a *woodpiece* or core, and
each woodpiece was sampled from a *tree* of a certain *species* at a
specific *site*. Typically, (some of) this information is encoded in the
name of each image. [Fonti et al.
(2025)](https://doi.org/10.3389/fpls.2016.00781) recommend the following
naming convention for QWA images and corresponding ROXAS output files:

`{site}_{species}_{tree/woodpiece}_{slide}_{image}`

e.g. `S22_LADE_L08b_1_2.jpg` would designate the *2nd* image of the
*first* slide from woodpiece *b* of the Larix decidua tree *L08* at site *S22*,
while `S17_LADE_L09_5_1.jpg` represents the *first* image of the *5th*
slide from the only woodpiece sampled from of the Larix decidua tree *L09* at
site *S17*.

The function `extract_data_structure()` can be used to extract this
information from the file names using a regular expression (regex) with
[named
groups](https://stringr.tidyverse.org/articles/regular-expressions.html#grouping:~:text=named%20capture%20group).
Note that not all hierarchical components need to be present in the
pattern, as long as each image can be uniquely identified. For example,
if there is exactly one image per slide, the slide (or image) identifier
can be omitted from the pattern.

```{r strc, eval = FALSE}
# Example pattern where the file names follow the recommended convention
# `{site}_{species}_{tree}_{slide}_{image}`
# with only one sample (woodpiece) per tree, so no woodpiece identifier is included in the pattern
pattern <- "^(?<site>[:alnum:]+)_(?<species>[:alnum:]+)_(?<tree>[:alnum:].+)_(?<slide>[:alnum:]+)_(?<image>[:alnum:]+)$"
df_structure <- extract_data_structure(files, pattern)
```

**Important:** Check the resulting dataframe with `View(df_structure)`
to ensure that the hierarchy was extracted correctly.

Alternative: if your files do not follow this naming convention, you can
provide it manually.... (TODO)

### Collecting QWA metadata

Next, read all the available metadata from the ROXAS Settings files and
the images directly.

```{r rawmeta, eval = FALSE}
df_rxsmeta <- collect_metadata_from_files(df_structure,
                                          roxas_version='classic')
```

To provide the additional metadata required (e.g. site and tree
information, author details, ...) for the submission of a dataset, you
can use the interactive Shiny app which is part of this package.

Launch the app with the following command: Inputs: df_rxsmeta from file
or environment, or json file Outputs: json

```{r metaapp, eval = FALSE}
launch_metadata_app()
```

The app is also hosted on a WSL server and accessible via
[webapps.wsl.ch/QWAmeta](https://webapps.wsl.ch/QWAmeta/).

### Cleaning and validating QWA data

To load and preprocess the cell and ring measurements:

```{r rawdata, eval = FALSE}
# read raw cells/rings data
QWA_data <- collect_raw_data(df_structure)

# clean raw data
QWA_data <- validate_QWA_data(QWA_data, df_meta)

# remove outliers
# NOTE: ROXAS does some threshold based outlier checks and assigns these a
# a negative value. Here, we replace these outliers with NAs
QWA_data <- remove_outliers(QWA_data)

# complete cell measures
QWA_data <- complete_cell_measures(QWA_data)

# save the cleaned data to files
write.csv(QWA_data$cells,
          file.path(path_out, paste0(dataset_name, '_cells.csv')),
          row.names = FALSE)

write.csv(QWA_data$rings,
          file.path(path_out, paste0(dataset_name, '_rings.csv')),
          row.names = FALSE)

```

To check the coverage and add ring quality flags and setting years to
exclude:

```{r, eval = FALSE}
launch_coverage_app()

# TODO: add export functionality to coverage app
```
