---
title: "rxs2tria"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rxs2tria}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction to rxs2tria

**Note:** This package is currently under development and not yet stable
nor officially released.

`rxs2tria` preprocesses a folder of
[ROXAS](https://www.wsl.ch/en/services-and-products/software-websites-and-apps/roxas/)
output files into a validated `QWAdata` object that is ready for analysis or submission to the
[TRIA database](https://webapps.wsl.ch/tria).

## Installation

You can install the development version of `rxs2tria` from
[GitHub](https://github.com/tria-db/rxs2tria/) with:

```{r, eval = FALSE}
pak::pak("tria-db/rxs2tria")
# pak::pak("tria-db/rxs2tria@develop") for the develop branch
```

## Quick start

The fastest way to get started is to copy the workflow template to your
working directory and fill in your dataset-specific paths and settings:

```{r, eval = FALSE}
file.copy(system.file("templates/prepare_rxs_dataset.R",
                      package = "rxs2tria"), ".")
```

The rest of this vignette walks through each step of that template in
detail.

## Step 1: Locate ROXAS files

```{r setup, eval = FALSE}
library(rxs2tria)
```

Set the path to a folder containing the ROXAS output files of your QWA
dataset. Files may be organized in subdirectories.
[get_roxas_files()] checks that all four required file types are present
for each image and returns their full filepaths:

- ROXAS cells output files (`{IMAGEID}_Output_Cells.txt`)
- ROXAS rings output files (`{IMAGEID}_Output_Rings.txt`)
- ROXAS settings files (`{IMAGEID}_ROXAS_Settings.txt`)
- Corresponding JPEG image files (`{IMAGEID}.jpg` or `{IMAGEID}.jpeg`)

```{r files, eval = FALSE}
path_in <- './path/to/ROXAS_output_data'
files <- get_roxas_files(path_in)
```

## Step 2: Extract the data structure

QWA data are organized hierarchically: each analyzed *image* was taken
of a microsection/*slide*, each slide was cut from a *woodpiece*, and each
woodpiece was sampled from a *tree* of a certain *species* at a specific
*site*. Typically, some or all of this information is encoded in the
image file names. 
[Fonti et al. (2025)](https://doi.org/10.3389/fpls.2016.00781) recommend
the naming convention:

`{site}_{species}_{tree/woodpiece}_{slide}_{image}`

e.g. `S22_LADE_L08b_1_2.jpg` designates the *second* image of the *first*
slide from woodpiece *b* of the *Larix decidua* tree *L08* at site
*S22*, while `S17_LADE_L09_5_1.jpg` represents the first image of the
5th slide from the *Larix decidua* tree *L09* at site *S17* (single
woodpiece, thus no woodpiece identifier needed).

[extract_data_structure()] extracts this information using a regular
expression with 
[named groups](https://stringr.tidyverse.org/articles/regular-expressions.html):

```{r strc, eval = FALSE}
# Example pattern for the recommended naming convention
# {site}_{species}_{tree}_{slide}_{image}
# (one woodpiece per tree, so no woodpiece identifier in the pattern)
pattern <- "(?<site>[[:alnum:]]+)_(?<species>[[:alnum:]]+)_(?<tree>[[:alnum:]]+)_(?<slide>[[:alnum:]]+)_(?<image>[[:alnum:]]+)"
df_structure <- extract_data_structure(files, pattern)
```

**Important:** Inspect the result with `View(df_structure)` to confirm
that the hierarchy was extracted correctly before proceeding.

Not all hierarchical components need to be present in the pattern, as
long as each image can be uniquely identified. If all images belong to
the same site or species, these can be supplied via the `site_label` and
`species_code` arguments instead of being part of the pattern.

## Step 3: Collect metadata

### From the ROXAS files

Extract image EXIF data and ROXAS settings and combine them with the
data structure into a `QWAmetadata` object.

*NOTE:* The timestamps in the image EXIF data and the ROXAS settings files
depend on your locale and/or operating system. To avoid conversion errors,
hey are originally read as pure character strings, and you need to 
explicitly convert them to POSIXct with the appropriate format(s) and timezone.

```{r rawmeta, eval = FALSE}
df_images <- collect_image_info(df_structure$fname_image)
df_settings <- collect_settings_data(df_structure$fname_settings,
                                     roxas_version = "classic")
df_images$img_created_at <- lubridate::parse_date_time(
  df_images$img_created_at, 
  orders = "%Y:%m:%d %H:%M:%S", # common EXIF format
  tz = "UTC" # commonly used in EXIF tags
)
df_settings$rxs_created_at <- lubridate::parse_date_time(
  df_settings$rxs_created_at,
  orders = c("%d.%m.%Y %H:%M:%S", "%d/%m/%Y %H:%M"),
  tz = Sys.timezone()
)
rxs_meta <- combine_rxs_metadata(df_structure, df_images, df_settings)
rm(df_images, df_settings)

rxs_meta  # prints a summary of the metadata
```

`combine_rxs_metadata()` returns a `QWAmetadata` object. Its `$images`
component holds the per-image metadata (ROXAS file paths, image data,
and ROXAS settings). The optional components (`$dataset`, `$authors`,
`$sites`, `$trees`, ...) are empty at this stage and can be provided
separately for a full TRIA database submission.

### Additional metadata (for TRIA submission)

To provide the site-, tree-, and dataset-level metadata required for a
TRIA submission, use the interactive Shiny app:

```{r metaapp, eval = FALSE}
launch_metadata_app()
```

The app is also hosted at
[webapps.wsl.ch/QWAmeta](https://webapps.wsl.ch/QWAmeta/).

## Step 4: Read and clean the measurement data

Read the raw cell and ring measurements into a `QWAdata` object, then
apply the three cleaning steps in order:

```{r rawdata, eval = FALSE}
# Read raw cells and rings data from all ROXAS output files
QWA_data <- collect_raw_data(df_structure)

# 1. Replace ROXAS automatic outliers (flagged as negative values) with NA
QWA_data <- remove_outliers(QWA_data)

# 2. Compute derived cell measures (tca, dcwt, cwtall.adj, ...) and
#    estimate earlywood / latewood width (eww, lww) per ring
QWA_data <- complete_cell_measures(QWA_data)

# 3. Validate dating, check for CWT estimates, and flag problematic rings.
#    Also attaches rxs_meta to the QWAdata object as $metadata.
QWA_data <- validate_QWA_data(QWA_data, rxs_meta)

QWA_data  # prints an overview including ring flag counts
```

After `validate_QWA_data()`, the `QWA_data$rings` dataframe contains
flag columns describing ring quality issues:

- `incomplete_ring`: ring at the inner or outer image border where the
  true ring boundary is not visible; MRW is unreliable or unavailable.
- `missing_ring`: no cells detected; ring was added manually during
  cross-dating (typically a wedging ring).
- `duplicate_ring`: the same year appears in more than one image due to
  overlapping slides from the same woodpiece.
- `exclude_dupl`: `TRUE` for duplicate rings not selected as the
  preferred source (ranked by cell count).
- `exclude_issues`: convenience flag combining incomplete and/or missing
  rings, depending on the `exclude_mode` argument.

To visually inspect coverage and ring flags across the dataset:

```{r coverage, eval = FALSE}
launch_coverage_app()
```

## Step 5: Save the results

```{r save, eval = FALSE}
path_out     <- './path/to/output'
dataset_name <- 'my_dataset'

write.csv(QWA_data$cells,
          file.path(path_out, paste0(dataset_name, "_cells.csv")),
          row.names = FALSE)
write.csv(QWA_data$rings,
          file.path(path_out, paste0(dataset_name, "_rings.csv")),
          row.names = FALSE)
```
