---
title: "Step-by-step instructions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Step-by-step instructions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks through each step of the data preparation process, from
raw ROXAS / ROXAS AI output files to a submission-ready TRIA dataset.

```{r setup, eval = FALSE}
library(rxs2tria)
```

# Step 1: Locate ROXAS files

Set the path to a folder containing the ROXAS or ROXAS AI output files
of your QWA dataset. Files may be organised in subdirectories.
`get_roxas_files()` checks that all required files are present for
each image and returns a data frame of their full file paths.

For **ROXAS**, the required files per image are:

- Cells output (`{IMAGEID}_Output_Cells.txt`)
- Rings output (`{IMAGEID}_Output_Rings.txt`)
- Settings file (`{IMAGEID}_ROXAS_Settings.txt`)
- Image file (`{IMAGEID}.jpg` or `.jpeg`)

For **ROXAS AI**, the required files per image are:

- Cells table (`{IMAGEID}.cells_table.csv`)
- Rings table (`{IMAGEID}.rings_table.csv`)
- Metadata file (`{IMAGEID}.metadata.json`)
- Image file (`{IMAGEID}.jpg` or `.jpeg`)

```{r, eval = FALSE}
roxas_version <- "roxas" # or "roxas_ai"
path_in <- "path/to/ROXAS_data"
files <- get_roxas_files(path_in, roxas_version)
```


# Step 2: Extract the data structure

QWA data are organised hierarchically: each analysed *image* was taken
of a microsection/*slide*, each slide was cut from a *woodpiece* (e.g.
a core or a wedge), and each woodpiece was sampled from a *tree* of a 
certain *species* at a specific *site*. Typically, some or all of this 
information is encoded in the image file names.
[Fonti et al. (2025)](https://doi.org/10.3389/fpls.2025.1505389) recommend
the naming convention:

`{site}_{species}_{tree/woodpiece}_{slide}_{image}`

e.g. `S22_LADE_L08b_1_2.jpg` is the second image of the first slide
from woodpiece *b* of the *Larix decidua* tree *L08* at site *S22*.

`extract_data_structure()` extracts this hierarchy using a regular
expression with
[named groups](https://stringr.tidyverse.org/articles/regular-expressions.html).
Each named group (`site`, `species`, `tree`, `woodpiece`, `slide`,
`image`) that appears in the pattern is extracted and used to construct
the full hierarchical labels. 
Not all groups need to be in the pattern:
If, for example, all images belong to the same site and/or species, the
`site_label` and/or `species_code` arguments can be supplied instead of
including `site`/`species` groups in the pattern. Or if each tree was cored
exactly once, there may not be a specific woodpiece identifier, and it 
can be left out of the pattern.

Adapt the pattern to match your naming convention using named regex groups.
Supported group names: site, species, tree, woodpiece, slide, image.
Example below assumes the common convention
  `{site}_{species}_{tree}_{slide}_{image}`
with alphanumeric components and one woodpiece per tree (no separate woodpiece
identifier). If your naming includes a woodpiece identifier, you can add it
 (e.g. by including "(?<woodpiece>[[:alnum:]]+)" between the tree and slide groups).
See the documentation of `extract_data_structure()` for more information.

```{r, eval = FALSE}
pattern <- "(?<site>[[:alnum:]]+)_(?<species>[[:alnum:]]+)_(?<tree>[[:alnum:]]+)_(?<slide>[[:alnum:]]+)_(?<image>[[:alnum:]]+)"
df_structure <- extract_data_structure(files, pattern)

# or:
# pattern <- "(?<tree>[[:alnum:]]+)_(?<slide>[[:alnum:]]+)_(?<image>[[:alnum:]]+)"
# df_structure <- extract_data_structure(
#   files, pattern, site_label = "S22", species_code = "LADE")
```

**Always inspect the result** with `View(df_structure)` before
proceeding to confirm that all labels were extracted correctly and the
hierarchy is consistent.


# Step 3: Collect metadata

## From the ROXAS or ROXAS AI files

`collect_settings_data()` reads the per-image ROXAS settings (or ROXAS
AI metadata) and, for classic ROXAS, also extracts EXIF data from the
image files. The result is combined with the data structure from Step 2
into a `QWAimages` object using `build_QWAimages()`.

Passing the data structure directly, the ROXAS version is auto-detected
from the settings file names, and for classic ROXAS the image EXIF data is
read from the `fname_image` column.

```{r, eval = FALSE}
df_settings <- collect_settings_data(df_structure)
```

Alternatively, the settings (and, for classic ROXAS, image) file paths can be
supplied individually, and the version can be set explicitly:

```{r, eval = FALSE}
df_settings <- collect_settings_data(files_settings = df_structure$fname_settings,
                                     files_images = df_structure$fname_image,
                                     roxas_version = "roxas")
```

Datetime columns are read as raw character strings to avoid locale- and
timezone-dependent conversion errors. You need to convert them
explicitly to `POSIXct` with the format(s) and timezone appropriate for
your data. Make sure to address any conversion warnings before overwriting
the columns in `df_settings`.

```{r, eval = FALSE}
img_created_at_converted <- lubridate::parse_date_time(
  df_settings$img_created_at,
  orders = "%Y:%m:%d %H:%M:%S", # common EXIF format
  tz = "UTC" # commonly used in EXIF tags
)
df_settings$img_created_at <- img_created_at_converted

# for ROXAS only (ROXAS AI has standardized timestamps, so already converted)
if (roxas_version == "roxas") {
  settings_date_orders <- c("%d.%m.%Y %H:%M:%S", "%d/%m/%Y %H:%M") # adjust to your locale
  rxs_created_at_converted <- lubridate::parse_date_time(
    df_settings$rxs_created_at,
    orders = settings_date_orders,
    tz = Sys.timezone()
  )
  df_settings$rxs_created_at <- rxs_created_at_converted
}
```

Combine the extracted data into one object:
```{r, eval = FALSE}
rxs_images <- build_QWAimages(df_structure, df_settings)
rm(df_structure, df_settings)

rxs_images  # prints a summary of the image-level metadata
```

`rxs_images` is a `QWAimages` object: a tibble subclass holding the
per-image metadata (file paths, image EXIF data, and ROXAS settings),
with the `roxas_version` stored as an attribute. It can be saved to a
compressed CSV for later use:

```{r, eval = FALSE}
write_QWAimages(rxs_images, "path/to/output_data/example_dataset_QWAimages.csv.gz")
# rxs_images <- read_QWAimages("path/to/output_data/example_dataset_QWAimages.csv.gz")
```

## Additional metadata for TRIA submission (Step 3b)

To provide the site-, tree-, woodpiece-, slide-, and dataset-level
metadata required for a TRIA submission, use the interactive metadata
Shiny app. It takes a `QWAimages` object as its starting point and
guides you through filling in the required fields. The output is a full
`QWAmetadata` object, which is exported as a `.json` file. Partially
completed submissions can be reloaded and continued at any time.

```{r, eval = FALSE}
launch_metadata_app()

# After exporting and closing the app, optionally read the result back:
# QWA_meta <- read_QWAmetadata("path/to/output_data/example_dataset_QWAmetadata.json")
```

---

# Step 4: Read and clean the measurement data

`collect_raw_data()` reads the cell and ring measurement tables from all
ROXAS output files, combines them, and performs initial cleaning:
negative values and error codes are replaced with `NA`, the dating is
checked for gaps or future years, and a warning is issued if CWT
estimates appear to be missing (relevant for conifer data).

`complete_QWAdata()` then adds derived cell and ring measures (lumen
area, CWT-based density, earlywood/latewood widths, etc.) and computes
ring quality flags:

- `incomplete_ring`: ring at the inner or outer image border where the
  true ring boundary is not visible; MRW is unreliable or unavailable.
- `missing_ring`: no cells detected; ring was added manually during
  crossdating (typically a wedging ring).
- `duplicate_ring`: the same year appears in more than one image due to
  overlapping slides from the same woodpiece.
- `exclude_dupl`: `TRUE` for the duplicate ring(s) not selected as the
  preferred source (ranked by cell count, penalising incomplete and
  missing rings).
- `exclude_issues`: convenience flag combining incomplete and/or missing
  rings, controlled by `exclude_mode`.

The `exclude_mode` argument determines which ring issues are
pre-selected for exclusion: `"either"` (default) marks both incomplete
and missing rings; `"incomplete_only"` marks only incomplete rings
(useful when missing/wedging rings should still contribute to
chronologies).

```{r, eval = FALSE}
QWA_data <- collect_raw_data(rxs_images)
# or if you already loaded the full metadata from json after completing Step 3:
# QWA_data <- collect_raw_data(QWA_meta$images)

exclude_mode <- "either"
QWA_data <- complete_QWAdata(QWA_data, rxs_images, exclude_mode)

QWA_data  # prints an overview including ring flag counts
```

---

# Step 5: Save the QWA data

`write_QWAdata()` saves the `QWAdata` object to two (optionally
compressed) CSV files following the naming convention
`{dataset_name}_QWAcells.csv(.gz)` and
`{dataset_name}_QWArings.csv(.gz)`. Explicit file names can also
be provided (see `?write_QWAdata`).

```{r, eval = FALSE}
path_out <- "path/to/output_data"
dataset_name <- "example_dataset"
write_QWAdata(QWA_data, dir = path_out, dataset_name = dataset_name)

# QWA_data <- read_QWAdata(dir = path_out, dataset_name = dataset_name)
```

At this stage you have the minimal components required for a TRIA
submission: the `QWAmetadata` `.json` file (from Step 3b) and the two
`QWAdata` `.csv` files. The following steps are optional but recommended
for a more complete submission.

---

# Step 6: Assess the measurements data quality (optional)

## Compute radial profiles (Step 6a)

Cell measurements can be aggregated across the ring width into
position bins, producing a `QWAprofile` object. Two binning schemes are
available:

- **Sector profiles**: divide each ring into *n* bins of equal *relative*
  width (0–100% of ring width). Comparable across rings of different
  absolute widths.
- **Band profiles**: sliding window of fixed *absolute* width and step
  size (in µm). Meaningful when absolute radial position matters (e.g.
  density gradients in µm from the cambium).

Both support computing means and optional quantiles per bin:

```{r, eval = FALSE}
prf_sector <- calculate_sector_profiles(QWA_data,
                                        n_sectors = 5,
                                        sel_cell_params = c("la", "cwttan"),
                                        quant_probs = c(0.1, 0.5, 0.9))

# prf_band <- calculate_band_profiles(QWA_data,
#                                     bandwidth = 50, stepsize = 25,
#                                     sel_cell_params = c("la", "cwttan"),
#                                     quant_probs = c(0.1, 0.5, 0.9))

# Save and reload:
# write_QWAprofile(prf_sector, "path/to/output_data/example_dataset_QWAprofile_sector5.csv.gz")
# prf_sector <- read_QWAprofile("path/to/output_data/example_dataset_QWAprofile_sector5.csv.gz")
```

---

## Visualise measurement series and flag rings (Step 6b)

The flags Shiny app provides an interactive overview of the ring
measurements and allows you to annotate individual rings with data
quality flags (e.g. compression wood, cracks) and discrete feature
flags (e.g. blue rings, frost rings).

The app takes as input (loaded from file or from the current R
environment):

- The `$rings` component of a `QWAdata` object (required).
- A sector `QWAprofile` (optional) — plotted alongside the ring
  time series to aid visual inspection.
- A `QWAimages` object (optional) — used to display image-level
  comments and to allow opening image files directly from the app.

The output is the `QWAdata$rings` data frame with additional logical
columns for each flag. It can be saved to file and read back to update
the `QWAdata` object:

```{r, eval = FALSE}
launch_flags_app()

# To update QWA_data$rings with the edited output:
# QWA_data$rings <- read_QWAdata(
#   file_rings = "path/to/output_data/example_dataset_edited_QWArings.csv")$rings
# QWA_data$rings <- df_rings_edited  # or directly from a data frame in the env
# write_QWAdata(QWA_data, dir = path_out, dataset_name = dataset_name)
```

---

## Update re-analysed images (Step 6c)

If data processing reveals issues in the raw data and you re-analyse
one or more images with ROXAS or ROXAS AI, `update_QWAimages()` and
`update_QWAdata()` allow you to splice in the new data without
reprocessing the entire dataset.

1. Replace the output files in `path_in` with the newly generated files.
2. Update the settings/metadata for those images in the `QWAimages`
   object:

```{r, eval = FALSE}
rxs_images <- update_QWAimages(rxs_images,
  imgs_to_update = c("IMAGELABEL1", "IMAGELABEL2"),
  settings_date_orders = settings_date_orders) # same as defined in Step 3
# don't forget to update the component in the QWAmetadata object if necessary:
QWA_meta$images <- rxs_images
```

3. Update the `QWAdata` object. For the updated images only, the raw
   data files are re-read, derived measures are recomputed, and
   automatic flags are reset. By default, any manual edits made in the
   flags app are also reset (set `reset_manual_flags = FALSE` to
   preserve them):

```{r, eval = FALSE}
QWA_data <- update_QWAdata(QWA_data,
  imgs_to_update = c("IMAGELABEL1", "IMAGELABEL2"),
  meta = rxs_images,
  exclude_mode = exclude_mode) # same as defined in Step 4
```

---

# Step 7: Compile supplementary resources (optional)

A TRIA submission must include at minimum the `QWAmetadata` `.json` and
the `QWAdata` `.csv(.gz)` files. If you want to provide additional
(supplementary) files (original images, annotated images, raw ROXAS /
ROXAS AI output, reference series `.rwl` files, etc.) these should 
submitted in a single zip alongside a standalone resources manifest 
listing all the files in the zip.

Not all files may be elligible as supporting material for a TRIA dataset,
and it is important to track which file pertains to which image or slide,
etc. For this purpose, `compile_resources()` scans a directory (including
nested subfolders) and returns a table with one row per file it finds.
For each file it infers the `resource_type` from the file name and sets
the hierarchy level it belongs to (`linked_level`). Files are also
matched to their `linked_label` (e.g. the `image_label`) using a
[QWAimages] object, where possible.

```{r, eval = FALSE}
QWA_meta <- read_QWAmetadata("path/to/output_data/example_dataset_QWAmetadata.json")
suppl_res <- compile_resources("path/to/submission_files", rxs_images = QWA_meta$images)
```

Every file gets a `status` (`"ok"`, `"review"`, or `"ignore"`) and the
console reports whether the supplementary path is ready for submission, 
or which files still need attention. Review the table, add a `description`
for files where the `resource_type` could not be inferred, and complete 
any `linked_label` values that were not filled automatically. Then
write it to a CSV to submit alongside the resources zip:

```{r, eval = FALSE}
suppl_res
vroom::vroom_write(suppl_res, "path/to/output_data/example_dataset_resources.csv")
```

See `vignette("resources")` for more details on this step, including how
to `recompile_resources()` after making edits, and how to fun a final 
`check_supplementary()` once you are done.

# Step 8: Submit your data

Once you are ready to share your prepared data files with us, email
[tria@wsl.ch](mailto:tria@wsl.ch)---either with a link to the files on a
file transfer service of your choice, or asking for an invite link to
WSL's own one. See `vignette("submission")` for what to upload and what
happens to your files once we receive them.